%&plain

\def\oth{{\rm otherwise}}

\footline{\tenrm Copyright \copyright\ 2013 Kieran Colford\hfill\folio}

\centerline{\bf The Running Time of Solving the Towers of Hanoi Puzzle}
\centerline{By: Kieran Colford}
\centerline{Date: Sunday, October $27^{th}$, 2013}

\beginsection Introduction

The Towers of Hanoi is a well known puzzle game in which there are
three rods (or towers), on the first rod there are a number of disks
of varying size and the goal of the game is to move each disk onto the
last rod.  Except there is a catch, you can only move one disk at a
time and you can only a disk on to an empty rod or on top of another
disk that is larger than it.

There is a well known recursive solution to this puzzle which is where
you move all but the largest disk (which is also the one placed at the
bottom) to an intermediate tower, then move the largest disk to the
destination, and then move the remaining disks on to the destination.
This algorithm has a running time of $M_1(n)=2^{n}-1$ where $n$ is the
number of disks.

It is interesting to study the question of how much faster is it to
solve this puzzle when the number of towers increases to more than
three.

\beginsection Begin

Due to the rules of the game, we know that to move any stack of disks
we must first move some top portion of the stack to an intermediate.
Then we must move the remaining disks.  Finally, the disks that were
set aside must be brought to their destination.

\proclaim Important Fact.  This is a direct consequence of the rules
that the puzzle is based on, and all legal configurations must be
achieved by this method.

All in all, if we let $M_k(n)$ be the number of moves necessary to
solve the puzzle.  Then we know that $M_k(n)=M_{k-1}(i)+2M_k(n-i)$ for
some integer $i$ which makes $M_k(n)$ as small as possible.
Therefore,
$$M_k(n)=\left\{\matrix{1&n=1\cr2^n-1&k=1\cr
Min\{M_{k-1}(i)+2M_k(n-i);i\in\left[1,n\right)\}&\oth\cr}\right.$$

Where $k$ is the number of towers excluding the first and last (so
it's the number of empty towers) and $n$ is the number of disks.

Then, by expanding the $M_k(n - i)$ term, we find
$$M_k(n)=\left\{\matrix{1&n=1\cr 2^n-1&k=1\cr
\sum^{G-1}_{b=0}2^bM_{k-1}(g_b)&\oth\cr}\right.\eqno{(1)}$$

Where $G$ is the size of some $g$ which minimizes $M_k(n)$ and
satisfies
$$n=\sum^{G-1}_{b=0}g_b$$

\beginsection Proof for k equals 2

Then, for the case when $k=2$ we have
$$\eqalign{M_2(n)&=\sum^{G-1}_{b=0}2^b(2^{g_b}-1)\cr
&=\sum^{G-1}_{b=0}2^{b+g_b}-\sum^{G-1}_{b=0}2^b\cr
&=\sum^{G-1}_{b=0}2^{b+g_b}-2^G+1\cr}$$

Let $a$ be the largest integer such that $T_a \le n$, where $T_n$ is
the $n^{th}$ triangular number (or $T_n={1\over 2}n(n+1)$).  We can
then construct $g$ such that $g=\{a,\dots,c,c,\dots,1\}$, where
$c=n-T_a$ and $G=a+1$.  Thus we have
$$\eqalign{M_2(n)&=\sum^{a-c}_{b=0}2^a+\sum^{a}_{b=a-c+1}2^{a+1}\
-2^{a+1}+1\cr &=(a-c+1)2^a+c2^{a+1}-2^{a+1}+1\cr&=(a+c-1)2^a+1\cr}$$

We know that this is the smallest configuration for $g$ because
$M_2(n)$ has a big-O notation of $O(\sqrt{n}2^{\sqrt{n}})$ and any
other configuration of $g$ would lead to an exponent greater than
$\sqrt{n}$.  A more rigorous proof of this can be constructed with
calculus, but this is sufficient for now.

\beginsection General Proof

Now to move on the general case of $M_k(n)$ for all $k$.  In order to
guarantee that the solution to $M_k(n)$ yields the minimum possible
value, we can represent $n$ as a sum of perfect inputs to $M_k$.
Conveniently, (1) gives us a way of computing such a thing.  As seen in
the previous proof for the case when $k=2$, the ideal input for $M_2$
is some triangular number $T_a$ which gives us $M_2(T_a)=(a-1)2^a+1$.
We then get the ideal input for $M_3$ like this,
$$\sum^{a-1}_{b=0}2^bM_2(T_{a-b})=M_3(\sum^{a-1}_{b=0}T_{a-b})\eqno{(2)}$$

% function f
Thus, the ideal input for $M_3$ is some sum of consecutive triangular
numbers.  In order to generalize this further, let us create a
function $f_k(a)$ which will produce the ideal input to a given $M_k$
so that $M_k(f_k(a))$ is as small as possible.  The solution to $f_k$
is,
$$f_k(a)=\sum^a_{b=1}f_{k-1}(b)$$

If the reader notices the relationship between this and pascal's
triangle, they will see that consequently we have the following three
identities,
$$f_0(a)=1$$
$$f_k(a)={a+k-1\choose k}$$
$$f_{k+1}(a)+f_k(a+1)=f_{k+1}(a+1)$$

\proclaim Side Note.  The second identity is useful because it allows
us to compute $f_k(a)$ without having to compute it's neighbours.
This ``Random Access'' allows a programmer to solve the largest
integer a satisfying $f_k(a)\le n$ by binary search rather than by the
much slower linear search.

% general
We can then simplify equation (2) further, like this,
$$M_3(f_3(a))=\sum^{a-1}_{b=0}2^bM_2(f_2(a-b))=h_3(a)2^a-1$$ 

For some $h_3(a)$.  We then see that the general form of the perfect
solution to $M_k$ is,
$$M_k(f_k(a))=h_k(a)2^a+(-1)^k\eqno{(3)}$$

For some $h_k(a)$.  So in the process of finding the above solution,
one naturally runs into the following solution for $h_k(a)$,
$$h_k(a)=\left\{\matrix{1&k=1\cr\sum^a_{b=1}h_{k-1}(b)-(-1)^k&\oth\cr}\right.$$

If one observes the fact that $h_k(a)$ is similar in construction to
$f_k(a)$, $h_k(a)$ can be further simplified into an alternating
series dependant on $f_k(a)$,
$$h_k(a)=\sum^{k-1}_{b=0}(-1)^{k+1-b}f_b(a)\eqno{(**)}$$

% function h
This can be seen from,
$$\eqalign{h_{k+1}(a)&=\sum^a_{b=1}h_k(b)\ +(-1)^k\cr
&=\sum^a_{b=1}\sum^b_{c=1}h_{k-1}(c)\ -a(-1)^k+(-1)^k\cr
&=\cdots\cr}$$ 

As we expand the function $h$ more and more, it ``sheds'' off an
alternating series that is dependant on $f$ because of the nested sums
each term was passed through.  Finally, when $h$ has reached its edge
condition, we are left with a pure nested sum, which is exactly
$f_k(a)$ (note the fact that the edge condition for $h$ is $1$ while
the edge condition for $f$ is $0$).  After some minor tweaking the
above equation can be easily simplified into (**).

% composition
To solve $M_k(n)$ for any $n$, let $a$ be the largest integer which
satisfies $f_k(a)\le n$ and $w=n-f_k(a)$.  This gives
$M_k(n)=M_k(f_k(a)+w)$, and if we repeat the above process so that we
get $u$, the largest integer such that $f_{k-1}(u)\le w$ (note the
$k-1$), and $v=w-f_{k-1}(u)$.  Thus $M_k(n)=M_k(f_k(a)+f_{k-1}(u)+v)$.
This can be further simplified to
$$M_k(f_k(a)+f_{k-1}(u)+v)=M_k(f_k(a))+2^{a-u}(M_k(f_k(u))+M_{k-1}(f_{k-1}(u)+v))\eqno{(4)}$$

This occurs because as was seen in the proof for $k=2$,
$$\eqalignno{M_k(f_k(a))&=\sum^{a-1}_{b=0}2^bM_{k-1}(f_{k-1}(a-b))\cr
&=\sum^{a-1-u}_{b=0}2^bM_{k-1}(f_{k-1}(a-b))+\sum^{a-1}_{b=a-u}2^bM_{k-1}(f_{k-1}(a-b))\cr
}$$

Then when we add in the the overflow term, we have to account for the
penalty we incurred by increasing the size of $g$.
$$\eqalignno{M_k(f_k(a)+f_{k-1}(u))&=\sum^{a-1-u}_{b=0}2^bM_{k-1}(f_{k-1}(a-b))+2^{a-u}M_{k-1}(f_{k-1}(u))+\sum^{a-1}_{b=a-u}2^{b+1}M_{k-1}(f_{k-1}(a-b))\cr
&=M_k(f_k(a))+2^{a-u}(M_k(f_k(u))+M_{k-1}(f_{k-1}(u)))&(*)\cr}$$

\noindent Then, (*) can simply be translated into (4).

% solution
In order to compute $M_k$ for any input, we can apply (4) and (3) to
get a solution to $M_k(n)$ (provided the decomposition if $n$ into
$\sum^k_{b=1}f_b(a_b)$ is already known).  Thus,
$$\eqalign{M_k(f_k(a)+w)&=M_k(f_k(a)+f_{k-1}(u)+v)\cr
&=M_k(f_k(a))+2^{a-u}(M_k(f_k(u))+M_{k-1}(f_{k-1}(u)+v))\cr
&=(h_k(a)+h_k(u))2^a+(-1)^k+2^{a-u}(M_{k-1}(w)+(-1)^k)\cr}$$ 

Thus, it is trivial to compute $M_k(n)$ for all $n$ and all $k$ using
the simple method of using the tail-recursive,
$$\eqalign{M_k(f_k(a)+w)&=\left\{\matrix{h_k(a)2^a+(-1)^k&w=0\cr
(h_k(a)+h_k(u))2^a+(-1)^k+2^{a-u}(M_{k-1}(w)+(-1)^k)&\oth\cr}\right.\cr
&{\rm where\;}u=Max\{i;i\in {\cal Z}, f_{k-1}(i)\le w\}\cr}$$

\noindent{\bf Q.E.D.}

\vfill

The following is a program written in python v3.3.2 that implements
the above method.  For details, read the enclosed docstrings and
comments.  Specifically, see the docstring for method m for an idea of
how powerful the algorithm is.  The module passes all test cases that
can be performed in a reasonable time.  The Tester class can be edited
to perform additional tests if one wishes.

\eject

\input test.py

\bye
